\documentclass[12pt]{article}
\usepackage[utf8]{inputenc}

\setlength{\oddsidemargin}{0in}  %left margin position, reference is one inch
\setlength{\textwidth}{6.5in}    %width of text=8.5-1in-1in for margin
\setlength{\topmargin}{-0.5in}    %reference is at 1.5in, -.5in gives a start of about 1in from top
\setlength{\textheight}{9in}     %length of text=11in-1in-1in (top and bot. marg.) 
\newenvironment{wileykeywords}{\textsf{Keywords:}\hspace{\stretch{1}}}{\hspace{\stretch{1}}\rule{1ex}{1ex}}

%%%%%%%%
%  Packages
%%%%%%%%

\usepackage{graphicx}% Include figure files
%\usepackage{caption}
\usepackage{amsmath,amssymb}
\usepackage[version=4]{mhchem}
\usepackage{siunitx}
\usepackage{longtable,tabularx}
\setlength\LTleft{0pt} 
\usepackage{color}% Include colors for document elements
\usepackage{dcolumn}% Align table columns on decimal point
\usepackage{bm}% bold math
\usepackage[numbers,super,comma,sort&compress]{natbib}
\usepackage{xcolor}
\usepackage{listings}
\lstset{
 language=bash, 
 basicstyle=\ttfamily\small,
 numberstyle=\footnotesize,
 numbers=left,
 backgroundcolor=\color{gray!10},
 frame=single,
 tabsize=2,
 rulecolor=\color{black!30},
 title=\lstname,
 escapeinside={\%(*}{*)},
 breaklines=true,
 breakatwhitespace=true,
 framextopmargin=2pt,
 framexbottommargin=2pt,
 extendedchars=false,
 inputencoding=utf8,
 showstringspaces=false
 }
\usepackage{dirtree}
\usepackage{indentfirst}
\newcommand{\red}[1]{\color{red}{#1}}

%%%%%%%%
%  Document
%%%%%%%%


\title{pyPolyBuilder user's guide and tutorials}

\newcommand*\samethanks[1][\value{footnote}]{\footnotemark[#1]}
\author{
	Mayk C. Ramos\thanks{Instituto de Qu\'imica, Universidade Federal do Rio de Janeiro, Rio de Janeiro 21941-909, Brazil},
	Patrick K. Quoika\thanks{Institute for General, Inorganic and Theoretical Chemistry, Center for Molecular Biosciences Innsbruck (CMBI), University of Innsbruck, A-6020 Innsbruck, Austria}, 
	Vitor A. C. Horta\thanks{Insight Centre for Data Analytics at Dublin City University, Dublin, Ireland},
	Douglas M. Dias\thanks{Dept. of Electronics and Telecomm. Eng., State University of Rio de Janeiro, Rio de Janeiro 20550-900, Brazil},\\
%	Elan G. Costa\thanks{Dept. of Electronics and Telecomm. Eng., State 
%	University of Rio de Janeiro, Rio de Janeiro 20550-900, Brazil},
	Elan G. Costa\samethanks{},
%	Jorge L.M. do Amaral\thanks{Dept. of Electronics and Telecomm. Eng., State 
%University of Rio de Janeiro, Rio de Janeiro 20550-900, Brazil},
	Jorge L.M. do Amaral\samethanks{},
%	Luigi M. Ribeiro\thanks{Dept. of Electronics and Telecomm. Eng., State 
%University of Rio de Janeiro, Rio de Janeiro 20550-900, Brazil},\\
	Luigi M. Ribeiro\samethanks{},\\
%	Klaus R. Liedl\thanks{Institute for General, Inorganic and Theoretical 
%Chemistry, Center for Molecular Biosciences Innsbruck (CMBI), University of 
%Innsbruck, A-6020 Innsbruck, Austria}\  and 
	Klaus R. Liedl\samethanks[2]\  and 
	Bruno A. C. Horta\samethanks[1]
%	Bruno A. C. Horta\thanks{Instituto de Qu\'imica, Universidade Federal do 
%Rio de Janeiro, Rio de Janeiro 21941-909, Brazil}
}

\begin{document}
\maketitle

\begin{abstract}
\end{abstract}

\tableofcontents

\renewcommand{\baselinestretch}{1.5}
\normalsize

\section{General concepts of the software}

PyPolyBuilder is a building block (BB) based automated topology builder.
Hence, based on a few BB molecular topology files (MTF), it is able to build a larger and more complex MTF in addition of a suggestion of the geometry as input for another molecular dynamics (MD) software.
Currently the code supports gromacs and gromos format.

PyPolyBuilder has two modules of application. The dendrimer module and the network one {\red{(also referred to as the polymer module)}}. 

\subsection{Dendrimer module}
PyPolyBuilder have a optimized and easier to use module in order to generate dendrimers (star-like polymers).
Due to its symmetry, the user is able to define the core, intermediary and terminal blocks only.
The connectivity is pre-assumed according to the desired generations of the dendrimer.
%Due to its symmetry, the user is able to instead of defining how each block is connected to each other, one can simply define the core, intermediary and terminal blocks and the desired generation of the dendrimer.
The software is prepared to make the linkage list and generate the final MTF and a initial guess for the dendrimer geometry in vacuum.

\subsubsection{Input files}
\label{sec:DENDinputFiles}
As inputs to dendrimer module, there is required the BBs topologies, the dendrimer generation and a force field parameters list.
In topologies files, there are six available fields to pass information that define the building block structure, similar to gromacs topology files (some example cases will be better discussed further in tutorial descriptions).
They are: \texttt{[ moleculetype ]}, \texttt{[ atoms ]}, \texttt{[ bonds ]}, \texttt{[ angles ]}, \texttt{[ dihedrals ]} and \texttt{[ branches ]}.

The field [ moleculetype ] is where the building block name should be written.
\begin{lstlisting}
[ moleculetype ]
; Name    nrexcl
   C4       2
\end{lstlisting}
This field defines the name of the BB in pyPolyBuilder's object-oriented workflow.
The second column, nrexcl, defines the number of excluded neighbors when used in gromacs. 
However, pyPolyBuilder does not actually use this value. It was maintained here for completeness and to conserve the similarities to the gromacs .itp files.

In the field [ atoms ], each atom should be listed with their index number, type, residue number, residue name, charge group number, coulombic charge and mass, in that order. The required values are:

\begin{lstlisting}
[ atoms ]
;  nr  type  resnr  resid  atom  cgnr  charge    mass    total_charge
    1    CH2   1    CORE      C    1    0.250  14.02700
    2    CH2   1    CORE      C    2    0.250  14.02700
    3    NT    1    CORE      N    1   -0.750  14.00670
    4    NT    1    CORE      N    2   -0.750  14.00670
\end{lstlisting}

It's important to notice that, pyPolyBuilder has a feature that modifies this pattern of the inputed gromacs-like topolgy.
Some common force fields (such as gromos), represent an aromatic center only by its improper torsionals.
Hence, in order to correctly model aromatic molecules when using a force field from the gromos family, one should exclude interactions between 1-4 neighbors in aromatic center.
Those interactions may easily be excluded from the final MTF generated by pyPolyBuilder by adding one column with the flag `AR' after the mass column, as illustrated below:

\begin{lstlisting}
[ atoms ]
;  nr  type  resnr  resid  atom  cgnr  charge    mass    
    1    C    1    AROM      C    1   -0.1298  12.01100   AR
    2    H    1    AROM      H    1    0.1298   1.00800   AR
    3    C    1    AROM      C    1   -0.1298  12.01100   AR
    4    H    1    AROM      H    1    0.1298   1.00800   AR
\end{lstlisting}

The `AR' flag should only be used for the atoms which are part of a aromatic center and for those that are directly bonded to aromatic centers.

The [ bonds ] section should have the information about which atoms are forming
chemical bonds. We need to pass their index numbers and information about
the force field parameters:
\begin{lstlisting}
[ bonds ]
;  ai   aj  funct   c0         c1
    1    2    2   gb_21
    1    3    2   gb_21
    2    4    2   gb_21
\end{lstlisting}
Where $a_i$ and $a_j$ are the index number of atom i and j, respectively, involved in the chemical bond and g\_21 is the force field parameter as the input of the gromacs package.

We strongly suggest that the bonds parameter are defined in each BB file.
However, it can also be defined using the atom types in the file that parse the parameters list.
This input will be discussed subsequently.

The last field that we need to provide for each BB is the [ branches ].
Here, the atoms are defined which are used to form a bond between two building blocks to propagate the monomer.
\begin{lstlisting}
[ branches ]
;  donor   acceptor
       1	 0
       0	 4
\end{lstlisting}

Usually, we can define the donor as the atom that will be bound in the new BB since it is donating a connection and the acceptor as the atom that will be bound in the last BB since it may accept a connection.

The parameters list file need to have the force field information about bonds, angles and dihedrals that exist in the topology.
Some cases may require that the stereochemistry conformation is specified, in these cases the conformation field can be used. In the cases where the conformation field is not present, the  final structure will have two of them. For example:

\begin{lstlisting}
# BONDS
H       N       gb_2
H       NT      gb_2
   

# ANGLES
H       NT      CH2     ga_11
H       NT      H       ga_10


# IMPROPERS
CH2     C       O       N        gi_1
C       N       H       CH2      gi_1


# DIHEDRALS
CH2     CH2     C       O        gd_40
CH2     CH2     C       N        gd_40


# CONFORMATION
CH2     C       N       CH2     trans
CH2     C       N       H       cis  
\end{lstlisting}

Notice that by default, the parameters from this parameters list file will be directly copied to the MTF.
By default, pyPolyBuilder uses an internal default set of interaction parameters for simplicity.
The ``\texttt{\# CONFORMATION}'' field set if the dihedral should be cis or trans and it selects the parameter to be used within this internal pseudo-force field.
In some complex cases, it might be interesting to make pyPolyBuilder to use the actual force field parameters (it will be exemplified in some of the following tutorials), this is possible by using the --forcefield option with a path to the force field to be used (in gromacs format).
All possible options are discussed in following sections.

\subsubsection{Command line call and optional flags}

To call the software, the following command must be run in the command line:

\begin{lstlisting}
pyPolyBuilder --core=core.itp \
              --inter=inter.itp \
              --ter=ter.itp \
              --params=list_param.itp \
              --ngen=1 \
              --name=NAME \
              --output=output.itp \
              --gro=output.gro \
              --dendrimer
\end{lstlisting}

All the possible options will be discussed exhaustively below.
There are some options which are specific for dendrimer module and may not be used if the dendrimer module is not being used.
To make it clear, those options are:

\begin{itemize}
    \item \textbf{\texttt{--core} or \texttt{-c}} should define a relative or absolute path for the BB that will be used as core.
        
    \item \textbf{\texttt{--inter} or \texttt{-i}} should define a relative or absolute path for the BB that will be used as intermediary.
    
    \item \textbf{\texttt{--ter} or \texttt{-t}} should define a relative or absolute path for the BB that will be used as terminal.
    
    \item \textbf{\texttt{--ngen} or \texttt{-n}} should be an integer defining the generation number of the dendrimer that will be build using pyPolyBuilder.
    
    \item \textbf{\texttt{--dendrimer}} sets the module that will be called by pyPolyBuilder.
\end{itemize}

Those are the specific options for the dendrimer module.
A complete list with the definition and usage of each option will be provided in a further section.

\subsection{Network module}
Differently from dendrimer module, network module does not assume any symmetry for the molecule.
Hence, all connections between BBs need to be defined.
Nevertheless, this module is absolutely general and can be used to build any molecule.

Here, the valence or connectivity of the BBs are not specified.
Because of that, using the \texttt{[ branches ]} field as in dendrimer module could lead to a cumbersome syntax.
This issue was avoided by defining a new input file that will be referred to as connectivity file.
This file may be considered as a topology of topologies.
%Philosophically, this file will work as a topology file for topologies.
All valence and connectivity for each BB is defined in this file.

\subsubsection{Input files}
\label{sec:GENinputFiles}

The BB files themselves are organized very similarly than they are in dendrimer modules (see previous section).
For instance, the fields \texttt{[ moleculetype ]}, \texttt{[ atoms ]}, \texttt{[ bonds ]}, \texttt{[ angles ]}, and \texttt{[ dihedrals ]} should be defined exactly the same. 
While the \texttt{[ branches ]} is not necessary anymore since the connections will be defined in the connectivity file.

In respect to the connectivity file, it is organized as follow.
There are two needed fields: ``\texttt{\# [BUILDING BLOCKS]}'' and ``\texttt{\# [ CONNECTS ]}'' as illustrated below:

\begin{lstlisting}
#[ BUILDING BLOCKS ]
;BBn     name
 1      PNIPS
 2      PNIP
 3      PNIP
 4      PNIP
 5      PNIPE

#[ CONNECTS ]
;BBi   BBj   ai   aj
  1     2    2    1
  2     3    2    1
  3     4    2    1
  4     5    2    1
\end{lstlisting}

``\texttt{\# [BUILDING BLOCKS]}'' field receives a list of all monomers that will be used.
The names that should be used are the names of the topology file as defined in the \texttt{[ moleculetype ]} field within the BB file.
For instance, the BBs should have a field where its name is defined (see previous section):
\begin{lstlisting}
[ moleculetype ]
; Name    nrexcl
   PNIP       2
\end{lstlisting}
with the BBs defined, the next field define how they are connected.

``\texttt{\# [CONNECTS]}'' field receives 4 values.
The two first are the index number of the two BBs that are being connected (as defined into ``\texttt{\# [BUILDING BLOCKS]}'' field) followed by the index number of the atoms within each BB that are being used to make the chemical bound.
For instance, by defining the line \#11 in the example connectivity file above ("1   2   2    1"), pyPolyBuilder connects the atom index 2 from the first BB to the atom index 1 from the second BB.

\subsubsection{Command line call and optional flags}
With all the BB, parameters list and connectivity files ready, pyPolyBuilder should be called using a command line as follows:

\begin{lstlisting}
pypolybuilder --bbs=bb_PNIP-start.itp,bb_PNIP.itp,bb_PNIP-end.itp \
              --in=connect-4.in \
              --params=list_param.itp \
              --name=PNIPAM \
              --output=PNIPAM.itp \
              --gro=PNIPAM.gro \
              --polymer
\end{lstlisting}

All possible options will be described in detail in the next section.
For clarity, the specific options for network module will be quickly discussed here as well.
These are:
\begin{itemize}
    \item \textbf{\texttt{--bbs} or \texttt{-bbs}} should be a list with the path for all BBs separated by comma.
    
    \item \textbf{\texttt{--in} or \texttt{-in}} should define a path for the connectivity file.
    
    \item \textbf{\texttt{--polymer}} sets the module that will be called by pyPolybuilder.
\end{itemize}

\subsection{Command line options}
\label{sec:CommandLine}

PyPolyBuilder help can be called to check a list of all the possible options to be used and a quick description for them by using the following command line:

\begin{lstlisting}
pypolybuilder -h
\end{lstlisting}

A complete list can be consulted below:

\begin{itemize}
  \item \textbf{\texttt{--help} or \texttt{-h}}:\\
    Show pyPolyBuilder help, a quick description of each option as well as the name of the attribute settled by the option.
    
  \item \textbf{Mandatory dendrimer options}
\begin{itemize}
  \item \textbf{\texttt{--core} or \texttt{-c}}: path\\
    This option inputs a path to a file that will be used as the core block in dendrimer module.
  \item \textbf{\texttt{--inter} or \texttt{-i}}: path\\
    This option inputs a path to a file that will be used as the intermediary block in dendrimer module.
  \item \textbf{\texttt{--ter} or \texttt{-t}}: path\\
    This option inputs a path to a file that will be used as the terminal block in dendrimer module.
  \item \textbf{\texttt{--ngen} or \texttt{-n}}: integer\\
    It inputs an integer that will set the number of the generation of the dendrimer that will be built.
\end{itemize}

    Each block should be built as previously described (See Section \ref{sec:DENDinputFiles}).
    These options are only needed if the dendrimer module is being used.
    
  \item \textbf{Mandatory network options}
\begin{itemize}
  \item \textbf{\texttt{--bbs} or \texttt{-bbs}}: comma-separated list of paths\\
    It receives a list containing the path to each BB used.
  \item \textbf{\texttt{--in} or \texttt{-in}}: path\\
    This option receives the path for the connectivity file dictate the structure of the molecule to be built.
 \end{itemize}   
    
    Each block and the connectivity file should be built as described in Section \ref{sec:GENinputFiles}.
    These options are only needed if the network module is being used.
    
  \item \textbf{Other mandatory options}
\begin{itemize}
  \item \textbf{\texttt{--params} or \texttt{-l}}: path\\
    This option receives the path for the parameters list file that will be included in the generated MTF.
\end{itemize}

  \item \textbf{Output control}
\begin{itemize}
  \item \textbf{\texttt{--output} or \texttt{-o}}: string\\
    It is used to define the name of the MTF. The resulting .itp file that will be generated will be named accordingly to this option. Default: default.itp
  \item \textbf{\texttt{--gro} or \texttt{-g}}: string\\
    It is used to define the name of the coordinates file in which the molecular structure is. The name given to the .gro file is settled by this option. Default: default.gro
  \item \textbf{\texttt{--name} or \texttt{-name}}: string\\
    The topology name is the identifier defined into the \texttt{[ moleculetype ]} field within the MTF (see Sections \ref{sec:DENDinputFiles} and \ref{sec:GENinputFiles}). This option sets the name for the topology. Notice that this name is what the network module uses for refer to the topology in its connectivity file. Default: NEWTOP
\end{itemize}

  \item \textbf{Optimization control}
\begin{itemize}
  \item \textbf{\texttt{--forcefield} or \texttt{-ff}}: path\\
    Since pyPolyBuilder uses an internal pseudo-force field to deal with most of common cases, it might be necessary to include real force field parameters in the optimization steps. This options receives the path to get the parameters values from a force field directory in gromacs format.
  \item \textbf{\texttt{--ngenga} or \texttt{-ngenga}}: integer\\
    Set the number of used genetic algorithm (GA) generations. Default: 20
  \item \textbf{\texttt{--npop} or \texttt{-npop}}: integer\\
    Set the number of used GA population. Default: 25
  \item \textbf{\texttt{--nsteps} or \texttt{-nsteps}}: integer\\
    Set the number of max steps used in geometry optimization. Default: 200
  \item \textbf{\texttt{--stepLength} or \texttt{-stepLength}}: float\\
    Length of the step used in the geometry optimization. In pyPolyBuilder, the geometry optimization step is not an actual energy minimization since some terms are omitted and by setting the \texttt{--nskipLJ} options, for instance, the objective function may be changed during the run. However, this is an indicative of the distance that atoms are moved at each optimization step. Default: 0.0001
  \item \textbf{\texttt{--nskipLJ} or \texttt{-nskipLJ}}: integer\\
    To avoid initial bad contacts resulting from the stochastic process from GA, the evaluation of Lennard-Jonnes (LJ) interactions can be avoided in the few first optimization steps. This option sets the number of iterations for which the LJ interactions will not be computed. Default: 0
\end{itemize}

  \item \textbf{Log control}
\begin{itemize}
  \item \textbf{\texttt{--verbose} or \texttt{-v}}:\\
    Be verbose. Tells pyPolyBuilder to explicit, in screen, every action it is making.
  \item \textbf{\texttt{--debug}}:\\
    Print debug messages.
  \item \textbf{\texttt{--nogeom}}:\\
    Do not perform geometry optimization. It can be used in cases that one need a MTF only and not a guess for the geometry. Also useful in saving time in building molecules that require calling pyPolyBuilder more than once by not optimizing intermediary structures. Applications of this option are given in the following tutorials
\end{itemize}

\item \textbf{Building Mode}
\begin{itemize}
  \item \textbf{\texttt{--dendrimer} or \texttt{-dendrimer}}:\\
    Set building mode to Dendrimer.
  \item {\red{\textbf{\texttt{--polymer} or \texttt{-polymer}}:\\
    Set building mode to Polymer.}}
\end{itemize}

\item \textbf{Output Formats}
\begin{itemize}
  \item \textbf{\texttt{--gromacs} or \texttt{-gromacs}}:\\
    Set output format to GROMACS. (This is default.)
  \item \textbf{\texttt{--gromos} or \texttt{-gromos}}:\\
    Set output format to GROMOS.
\end{itemize}

\end{itemize}


\section{Tutorials}

This section is divided in three sections:
\begin{itemize}
    \item \textbf{Dendrimer module}
    where some examples using only the dendrimer module are provided;
    \item \textbf{Network module}
    where the tutorials discussed use only the network module; and
    \item \textbf{Advances examples}
    where both modules were combined to ease the construction of the molecule.
\end{itemize}

Each tutorial directory has a similar structure. 
The needed BBs are provided in the files with \texttt{.itp} extensions.
In particular, BBs for dendrimers are named according to its topology: \texttt{core\_**.itp}, \texttt{inter\_**.itp} and \texttt{ter\_**.itp} for the core, intermediary and terminal BBs.
In addition, a bash script named \texttt{how\_to\_run\_this\_example.txt} can be used to automatically run the tutorial.

After using pyPolyBuilder to generate the geometry and the MTF, these files can be tested in the run directory.
This directory has the needed files to carry out a MD simulation using gromacs.
In order to run the simulations, it is needed to copy the output from pyPolyBuilder in the run directory.
That is, if the PAMAM tutorial is being studied, as example, after running the pyPolyBuilder, the following command line needs to be used:

\begin{lstlisting}
cp PAMAM.* run/.
\end{lstlisting}

In the run directory, there is a bash script with \texttt{.sh} extension and named in accordance with the tutorial name.
This file can be used to automatically run a workflow for solvate, equilibrate and a MD simulation steps.
The equilibration step is divided in two steps, a nvt equilibration for 100 ps and also a npt for 100 ps.
By default, the MD simulation step will run 100 ps of production too, for time saving purposes.
Before executing the bash file, it needs to be edit to use the correct gromacs path on you machine.

\subsection{Dendrimer module}
In the dendrimer module, pyPolyBuilder is automated to build a dendrimer based on three BBs.
The force field parameters will be provided in the software through a file defined used the \texttt{--params} option.
After the building blocks are defined, the only needed input is the dendrimer generation that should be passed through the \texttt{--ngen} option.
PyPolyBuilder is programmed to calculate the number of needed monomers to generate a dendrimer of the desired generation number and to connect all of them in order to output a final topology file and an initial guess for a coordination file that should be minimized afterwards with a proper MD package.

Here we provide some simple tutorials aiming to give the user some prototype files and to pass the philosophy of using the dendrimer module of pyPolyBuilder.

\subsubsection{PAMAM}
\input{PAMAM/PAMAM}
\clearpage

\subsubsection{PPI}
\input{PPI/PPI}
\clearpage

\subsubsection{SPL7013}
\input{SPL7013/SPL7013}
\clearpage

\subsubsection{PAMAM/PPI-Janus}
\label{PAMAM/PPI-Janus}
\input{PAMAM_PPI-Janus/PAMAM_PPI-Janus}
\clearpage

\subsection{Network module}
Differently from the dendrimer module, there is no pre-assumed polymer topology.
Hence, the user needs to completely define how every BB is connected.
In order to make all information about the molecule connectivity available in one single file, the network module does not use the \texttt{[ branches ]} field.
Besides, all the connections should be passed in the connectivity file through the \texttt{--in} option.
Its content was already discussed in Section \ref{sec:GENinputFiles}.
Also, when using this module, all the building blocks should be passed using the \texttt{--bbs} option (differently from dendrimer module that uses the options \texttt{--core}, \texttt{--inter} and \texttt{--ter}).

Here we provide some simple tutorials aiming to give the user some prototype files and to pass the philosophy of using the network module of pyPolyBuilder.

\subsubsection{PEG}
\input{PEG/PEG}
\clearpage

\subsubsection{PNIPAM}
\input{PNIPAM/PNIPAM}
\clearpage

% \subsubsection{poly_p-benzamide}
% \input{Poly_Benzamide/PBZ}

% \subsubsection{Cyclic polymer}
% \input{Cyclic_Pol/Cyclic_Pol}


\subsection{Advanced examples}
For some complex cases, it may be necessary to combine both modules to build different parts of the molecule separately. 
In the following tutorials parts of the molecule were built using the dendrimer module and the network module was used to connect them.
Generally, these tasks may also be realized with the network module alone since it is completely generic, however, the dendrimer module is specialized to build any dendritic structure and it is desirable to use it to ease the process.

\subsubsection{PAMAM/PPI-half}
\input{PAMAM_PPI-half/PAMAM_PPI-half}
\clearpage

\subsubsection{PEG connected PAMAM dendrimer}
\input{PAMAM_PEG/PAMAM_PEG}
\clearpage

% \section*{Appendix}


\end{document}
